Section Three

### 2020-10-13 new2020/sections/sec03.Rmd
pf prediction steps

**Step 2: compute arrival time at next stop**

If there are any remaining segments between the current segment $\TripSeg_k\vi$ and the next stop $j_k\vi + 1$, their travel times are sampled from the network state with variance increased by network system noise to account for forecasting changes.
\[
   v_j\vi \sim
   \TNormal{\hat \NWstate_j}{\left[(\RouteNWstatevarseg_j + \ttsofar\vi\NWnoise)^2 + \NWvar_j^2\right]\wedge P_{\text{max}}}{0}{\MaxSpeed_j},
\]
allowing uncertainty to increase up to a maximum $P_\text{max}$ which is the overall variance of all speeds along all segments estimated from historical data. The travel time along each segment is added to the travel-time-so-far,
\[
   \ttsofar\vi = \ttsofar\vi + \frac{\Tseglen_j}{v_j\vi}.
\]

Now, arrival time at stop $j$ is the sum of travel time along all segments up to the next stop. In the iterative process, this is simply $\Tarr_j\vi = t_k + \ttsofar\vi$.

**Step 3: compute stop dwell time**

Once at a stop, a bus either stops and waits while passengers board and disembark, or travels past without stopping. This *dwell time* needs to be modelled @citep:cn;, but adds multimodality to the arrival time distribution.

The dwell time model is ... [from chapter 3], as used by @citet:Hans_2015,Elliott_2020;. This now gets added to the travel-time-so-far,
\[
   \ttsofar\vi = \ttsofar\vi + \pdwell_j\vi
\]
unless the stop is a layover. In that case, we allow the particle---with some probability---to remain at the stop until the end of the dwell time, or the scheduled departure time, whichever is later.

**Step 4: repeat steps 2--3**

We now repeat each of the steps, sampling travel times for intermediary road segments as the particle arrives, and storing stop arrival times, until the particle reaches the end of the route.

**Step 5: obtain arrival time distributions**

Since each particle stores its arrival at all stops, we can very easily obtain the predictive distribution of arrival time for the bus using the delta measure,
\[
   p(\Tarr_j | \Vstate_k, \vec\NWstate_k) \approx
   \frac{1}{\tilde N}\sum_{i=1}^{\tilde N}
      \DiracMeasure{\Tarr_j\vi}{\Tarr_j}.
\]
We may obtain any summary statistics from this distribution, including the mean and prediction intervals.

>>>>

#### **Step 2: compute arrival time at next stop**

If the next node is an intersection, as is the case in [[figure X]], we iteratively calculate travel time along each segment until the particle arrives at a stop. As before we simulate vehicle speed from the network state, but now allowing for increased variability over time. Here, a forecast function could be used if available, otherwise a simpler approach can be used, as we have here. We use a similar distribution as in step\ 1, but allow speed to increase over time with a rate of $q$\ m/s$^2$, up to a maximum of $P_\text{max}$, which is the overall variance of vehicle speeds based on historical data. The predicted particle speed is thus
\[
   v_m\vi \sim
   \TNormal{\hat \NWstate_m}{\left[(\NWstatevar_{mm} + \ttsofar\vi\NWnoise)^2 + \NWvar_m^2\right]\wedge P_{\text{max}}}{0}{\MaxSpeed_m}.
\]
Travel-time-so-far is increased by the particle's travel time,
\[
   \ttsofar\vi = \ttsofar\vi + \frac{\Tseglen_m}{v_m\vi}.
\]

This process is repeated until the particle arrives at a stop, as is indicated by III in [[figureX]]. The arrival time at the stop is then stored as
\[
\Tarr_m = t_k + \ttsofar\vi.
\]

#### **Step 3: stop dwell time**

When at a stop, the particle first determines its dwell time. Using the model parameters described in \cref{particle-filters-estimating-vehicle-state}, the dwell time at stop $j$ is added to the travel-time-so-far,
\begin{equation*}
\begin{split}
   p_j\vi &\sim \Bern{\pi_j},\\
   d_j\vi &\sim \TNormal{\dwell_j}{\dwellvar_j}{0}{\MaxSpeed_j},\\
   \ttsofar\vi &= \ttsofar\vi + p_j\vi(\gamma + d_j\vi).
\end{split}
\end{equation*}

However, we also need to consider *layovers*, which are stops at which the bus is expected to wait if it is running early. Adherence is shown in [[figureY]], which shows about 60% driver compliance. So, with probability 0.6, the particle will wait until at least the scheduled departure time at stop $j$, $D_j$:
\[
   \ttsofar\vi =
   \begin{cases}
      \max\{\ttsofar\vi, D_j - t_k\} & \text{w.p. } 0.6,\\
      \ttsofar\vi & \text{otherwise}.
   \end{cases}
\]

<!-- Once at a stop, a bus either stops and waits while passengers board and disembark, or travels past without stopping. This *dwell time* needs to be modelled @citep:cn;, but adds multimodality to the arrival time distribution.

The dwell time model is ... [from chapter 3], as used by @citet:Hans_2015,Elliott_2020;. This now gets added to the travel-time-so-far,
\[
   \ttsofar\vi = \ttsofar\vi + \pdwell_j\vi
\]
unless the stop is a layover. In that case, we allow the particle---with some probability---to remain at the stop until the end of the dwell time, or the scheduled departure time, whichever is later. -->

#### **Step 4: repeat steps 2--3**

Next, the particle continues along the route, sampling travel times (step 2) for each road segment, and dwell times (step 3) for each stop, while recording its arrival time at each stop along the way until it reaches the end of the route.

#### **Step 5: obtain arrival time distributions**

The predictive distribution of arrival time at stop $j$, given trip and network states at time $t_k$, is easily obtained using the delta measure (section XX),
\[
   p(\Tarr_j | \Tripr_k, \vec\NWstate_k) \approx
   \frac{1}{\tilde N}\sum_{i=1}^{\tilde N}
      \DiracMeasure{\Tarr_j\vi}{\Tarr_j}.
\]
Means, quantiles, and other summary statistics can easily be obtained from the sample; however, some, such as quantiles, require sorting all $\tilde N$ particles. In \cref{chapter4} I present an approximation which avoids sorting the particles which is a computationally intensive procedure.
